Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/elife.tex
Better fig name: 3 -> ablation

\begin{figure}
    \includegraphics[width=\linewidth]{fig3}
    \caption{%
        Deviance detection under model ablation.
        \textbf{A} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel C). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
        \textbf{B} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics.
    }
    \label{fig:3}
\end{figure}

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{3}A shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{3}C) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

Finally, based on the ablated model analysis in \FIG{threshold-initial}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}F, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}G, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{3}.

>>>>

\begin{figure}
    \includegraphics[width=\linewidth]{fig-ablation}
    \caption{%
        Deviance detection under model ablation.
        \textbf{A} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel C). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
        \textbf{B} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics.
    }
    \label{fig:ablation}
\end{figure}

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{ablation}A shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{ablation}C) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

Finally, based on the ablated model analysis in \FIG{threshold-initial}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}F, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}G, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{ablation}.

### 2023-03-20 paper-1/manuscript.tex
Better fig name: 3 -> ablation

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig:3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation.
\textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
\textbf{b} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{3}a shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{3}c) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

Finally, based on the ablated model analysis in \FIG{threshold-initial}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}f, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}g, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{3}.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig-ablation.pdf}
    \caption{}
    \label{fig:ablation}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation.
\textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA).
\textbf{b} Deviance detection indices across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. \FIG{ablation}a shows the responses to target stimulation of the sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the \dev{} sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high \dev{}, intermediate \msc{}, and low \std{} responses, yielding a DDI of 0.19. Finally, in the full model, the \dev{} response stood out more clearly, leaving much smaller \msc{} and \std{} responses and yielding a DDI of 0.4.

Finally, across networks and stimuli, the DDI (\FIG{ablation}c) was exactly 0 as expected in the ablation with no plasticity, not significantly different from 0 with STD only (T = 3.2e+03, p = 0.262, median -2.1e-04; two-sided Wilcoxon, n = 120), but clearly positive for TA only (T = 5.85e+03, p = 3.1e-09, median 0.077; one-sided Wilcoxon, n = 120) and the full model (T = 5.97e+03, p = 4.75e-10, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: T = 6.1e+03, p = 4.95e-11; TA: T = 4.87e+03, p = 0.000604), and greater too than the sum of the DDI values across both STD only and TA only models (T = 5.12e+03, p = 4.52e-05). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

@@

Finally, based on the ablated model analysis in \FIG{threshold-initial}, we sought to confirm that lowered $\T{A}{}$ was responsible for the increased target response. As shown in \FIG{addingSTD}f, $\Delta \T{A}{exc}$ and $\Delta \R{A}{exc}$ were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response remained clearly increased in \dev{} over \msc{}, as shown in \FIG{addingSTD}g, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians of $\Delta \R{A}{}$ differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in \FIG{ablation}.
